---
title: "STA3100 Final Project Report"
author: "Anjali Tomerlin"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_title: "Table of Contents"
    toc_float: true
    theme: cosmo
date: "2025-11-03"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("data_scraping.R")
library(ggplot2)
```

```{r}
lst <- get_work_counts()

#basic data investigation
ggplot(lst$Mathematics,aes(x=year,y=counts)) + geom_point() + labs(title="Mathematics")
ggplot(lst$Biology,aes(x=year,y=counts)) + geom_point() + labs(title="Biology")
ggplot(lst$History,aes(x=year,y=counts)) + geom_point() + labs(title="History")
ggplot(lst$CS,aes(x=year,y=counts)) + geom_point() + labs(title="Computer Science")
ggplot(lst$Medicine,aes(x=year,y=counts)) + geom_point() + labs(title="Medicine")
ggplot(lst$Art,aes(x=year,y=counts)) + geom_point() + labs(title="Art")
ggplot(lst$Religion,aes(x=year,y=counts)) + geom_point() + labs(title="Religion")
ggplot(lst$Philosophy,aes(x=year,y=counts)) + geom_point() + labs(title="Philosophy")
ggplot(lst$Literature,aes(x=year,y=counts)) + geom_point() + labs(title="Literature")
```


# Abstract

# Introduction

## Background Information

Immediately following the advent of the COVID-19 pandemic in March of 2020, the focus of much academic research and funding was pivoted towards medicine. However, the long-term impact of this shift on the research in other areas, particularly the humanities, has often been overlooked. 

Five years after the initial COVID-19 outbreak, we aim to investigate trends in the number of published works in humanities-related fields---including history, philosophy, art, literature, and religion---before and after the pandemic. We also aim to compare these emerging trends with those in medicine-related research. 

# Data

# Analysis and Result 

# Conclusion

# Works Cited

# Contribution Statement



